### 2014-03-27 text/Chapter04.tex
Finish introduction chapter and restructure the rest

\chapter{Structural coeffect language} 
\label{ch:structural-coeffects}

\newcommand{\ctimes}{ \hat{\times} }
\newcommand{\cdupl}[0]{ \Delta } %{ \Upsilon }

\newcommand{\csprd}[0]{ \times }
\newcommand{\csvop}[0]{ \vee }

% alculus names and rule names
\newcommand{\clflt}{$\lambda_{\textnormal{\sffamily Cf}}$}
\newcommand{\clstr}{$\lambda_{\textnormal{\sffamily Cs}}$}

\newcommand{\ruleft}[1]{\textsf{\clflt\small{-T#1}}}
\newcommand{\rulefs}[1]{\textsf{\clflt\small{-S#1}}}
\newcommand{\rulest}[1]{\textsf{\clstr\small{-T#1}}}
\newcommand{\ruless}[1]{\textsf{\clstr\small{-S#1}}}
\newcommand{\rulessc}[1]{\textsf{\clstr\small{-SC#1}}}

@@

The \emph{flat coeffect system} presented in the previous sections has a number of uses, but often
we need to track context-dependence in a more fine-grained way. To track neededness or security, 
we need to associate information with individual \emph{variables} of the context.

%===================================================================================================

\section{Introduction}

@@

\subsection{Motivation: Tracking array accesses}
\label{sec:structural-motivation}
%\newcommand{\pastval}[2]{\ident{#1}_{[#2]}}
%\newcommand{\dnat}{\ident{D}_\ident{nat}}

Similarly to the flat version, the \emph{structural coeffect calculus} works with contexts and
functions annotated with a coeffectt tags, written $\ctyp{r}{\Gamma}$ and $\ctyp{r}{\tau_1} \rightarrow \tau_2$, 
respectively, but we use richer tag structure.

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\csprd$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \csprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The coeffect tag $5 \csprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
a single variable:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \csvop 10}{(\ident{c}:\dnat)}
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
In this example, the operation $\csvop$ would be the \emph{max} function and so $5 \csvop 10 = 10$.
Before looking at the formal definition, consider the typing of let bindings:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
\pastval{a}{15}+\pastval{c}{10}
\end{array}
\end{equation*}
%
The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.

The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
context is $\ctyp{0 \csprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
of $\csvop$ and it denotes empty coeffect. The typing context of the
body is $\ctyp{15 \csprd 10}{(\ident{a}, \ident{c})}$.

To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
of the context in which \ident{c} was defined using the $\csvop$ operation. This is then combined 
with the remaining tags from the body yielding the overall context:
$\ctyp{15 \csprd (10 \csvop (0 \csprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
normalization mechanism (described later), this can be further reduced to
$\ctyp{(15 \csvop 10) \csprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
\ident{b}.

@@

\subsection{Structural coeffect tags}
In the previous section, the $\csvop$ operation behaves similarly to the flat $\vee$ operation.
However, the type system does not require some of the semilattice properties, because some uses
are replaced with the $\csprd$ operation. We do not require any properties about the $\csprd$ operation.
For example, in the previous example cannot be commutative (since a tag $15 \csprd 10$ has different 
meaning than $10 \csprd 15$). However, we relate the operations using distributivity laws to allow 
normalization that was hinted above.

\begin{definition}
A \emph{structural coeffect tag structure} $(S, \csprd, \csvop, 0, 1)$ is a tuple where 
$(S, \csvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
by a binary operation $\csprd$ and element $1\in S$ such that
for all $r,s,t\in S$, the following equalities hold: 
%
\begin{subequations}
\begin{align}
\tag{associativity}
r \csvop (s \csvop t) &= (r \csvop s) \csvop t\\
\tag{commutativity}
r \csvop s &= s \csvop r\\
\tag{lower bound}
r \csvop 0 &= r\\
\tag{distributivity}
r \csvop (s \csprd t) &= (r \csvop s) \csprd (r \csvop t)\\
\tag{upper bound}
1 \csvop r &= 1
\end{align}
\end{subequations}
%
\end{definition}
%
% 'upper bound' (not sure what to call it) says that empty context cannot carry information
% looks like this is the upper bound of the bounded lattice ???
%
% commutativity is not strictly required by any law or type preservation, but
% it simplifies the presentation and looks reasonable
%
The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
always accessed using standard variable access). The tag $1$ is used to annotate empty 
variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
to carry an annotation. We explain exactly how this works when we introduce the type system.
The fact that $1$ is the upper bound means that combining it with other coeffect annotations
does not affect it and so empty contexts cannot carry any information.

The structure generalizes the \emph{flat coeffect tag structure} introduced in 
Section~\ref{sec:flat-tag-structure}, but it additionally requires a special element $1$ 
representing the upper bound. Given a flat coeffect structure, we can construct a structural
coeffect structure (but not the other way round). For certain structures, the element $1$
may be already present, but in general, it can be added as a new element.
This construction will be important in Section~\ref{sec:generalization},
where we show that \clstr~calculus generalizes~\clflt.

\begin{lemma}
\label{thm:structural-generalizaton}
A flat coeffect tag structure $(S, \vee, 0)$ implies a structural coeffect tag structure.
\end{lemma}
\begin{proof}
Take $1 \notin S$, then $(S \cup \{1\}, \vee, \vee, 0, 1)$ is a structural coeffect tag 
structure. From the properties of flat coeffect tag structure, we get that the $\vee$ operation 
is associative, commutative and $0$ is the unit with respect to $\vee$.

To prove the distributivity, we need to show $r \vee (s \vee t) = (r \vee s) \vee (r \vee t)$.
This easily follows from commutativity, associativity and idempotence of the flat coeffect tag 
structure.
\end{proof}

@@

\subsection{Structural coeffect type system}
\label{sec:structural-types}

The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
a \emph{structural coeffect tag structure} $(S,\csprd, \csvop, 0)$.

The system differs from \clflt~in a significant aspect. It contains explicit 
structural rules that manipulate with the context. Such rules allow reordering, duplicating
and other manipulations with variable context. Such rules are known from affine or linear
type systems where they are removed to obtain more restrictive system. In our system, the
rules are present, but they manipulate the variable structure $\Gamma$ as well as the 
associated tag structure $r$.

As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
simple set. Instead, we use the following tree-like structure (which is more similar to 
bunched types than to linear or affine systems):
%
\begin{equation*}
\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
\end{equation*}
%
The syntax $()$ represents an empty context, so the structure defines a binary trees where
leaves are either variables or empty. Contexts such as $\ctyp{1\csprd(r \csprd 1)}{((), (x, ()))}$ 
contain unnecessary number of empty contexts $()$. However, we need to construct them
temporarily, because certain rules require splitting a context and, by our 
definition, the context $(x:\tau)$ is not splitable.

The typing rules of the system are shown in Figure~\ref{fig:structural-coeffect-typing}.
Many of the structural rules are expressed in terms of a helper judgement
$\ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{r}{\Gamma}}$.

\paragraph{Standard rules.}
Variable access (\rulest{var}) annotates the corresponding variable with an empty coeffect
$0$. The \rulest{fun} rule assumes that the context of the body can be split into the
variable of the function and other (potentially empty) context and it attaches the coeffect
associated with the function variable to the resulting function type $\ctyp{s}{\tau_1} \rightarrow \tau_2$.

The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
expression, argument and the function type respectively. 
The result of evaluating the argument in the context $t$ is passed to the function that
requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
the combination of coeffects $r \csvop t$. The variable context required to evaluate the 
function value is independent and so it is annotated just with coeffects $s$.
Finally, the \rulest{let} rule is derived from let binding and application, but we show
it separately to aid the understanding.

\paragraph{Structural rules.}
The remaining rules are not syntax-directed. They are embedded using the \rulest{ctx}
rule and expressed using a helper judgement $\ctxtrans{\ctyp{r_1}{\Gamma_1}}{\ctyp{r_2}{\Gamma_2}}$
that says that the context on the left-hand side can be transformed to the context on the
right-hand side. The transformation can be applied to any part of the context, which 
is captured using the \rulest{nest} rule (it is sufficient to apply the transformation 
on the left part of the context; the right part can be transformed using \rulest{exch}).

The \rulest{empty} rule allows attaching empty context to any existing context. The rule 
is needed to type-check lambda abstractions that do not capture any outer variable.
The \rulest{weak} rule is similar, but it represents \emph{weakening} where an unused variable is 
added. The associated coeffect tag does not associate context information with the variable.
This is needed to type-check lambda abstraction that does not use the argument.

The \rulest{contr} rule is a limited form of contraction. When a variable appears repeatedly,
it can be reduced to a single occurrence. The rule is needed to satisfy the side condition
of \rulest{fun} when the body uses the argument repeatedly. The two associativity rules
together with \rulest{exch} provide ways to rearrange variables. Finally, \rulest{sub} represents
sub-coeffecting -- in \clstr~the rule operates on coeffects of individual variables.

\ExecuteMetaData[rules/rules.tex]{scl-types}

@@

\subsection{Properties of reductions}
\label{sec:structural-metatheory}
Similarly to the flat version, the \clstr~calculus is defined abstractly. We cannot define its
operational meaning, because that will differ for every concrete application. For example,
when tracking array accesses, variables are interpreted as arrays and $\pastval{a}{n}$ denotes
access to a specified element.

Just like previously, we can state general properties of the reductions. As the syntax of expressions
is the same for \clstr~as for \clflt, the substitution and reduction $\twoheadrightarrow_\beta$ 
are also the same and can be found in Figure~\ref{fig:flat-substred}.

The structural coeffect calculus \clstr~associates information with individual variables. This means
that when an expression requires certain context, we know from what scope it comes -- the
context must be provided by a scope that defines the associated variable, which is either a
lambda abstraction or global scope. This distinguishes the structural system from the flat system
where context could have been provided by any scope and the lambda rule allowed arbitrary 
splitting of context requirements between the two scopes (or declaration and caller site).

\paragraph{Internalized substitution.}
Before looking at properties of the evaluation, we consider let binding, which can be 
viewed as internalized substitution. The typing rule \rulest{let} can be derived from 
application and abstraction as follows.

\begin{lemma}[Definition of let binding]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
\end{lemma}
\begin{proof}
The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
correspond with the typing rule \rulest{let}:
\begin{equation*}
\inference[]
  {\inference
    {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
\end{equation*}
\end{proof}

The term $e_2$ which is substituted in $e_1$ is checked in a different variable
and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
the variable that we are substituting for. The let binding captures substitution for a 
specific variable (the context is of a form $\ctyp{r \csprd s}{\Gamma, v:\tau}$).
For a general substitution, we need to define the notion of context with a hole.

\newcommand{\chole}[4]{\ctyp{#1}{#2}[#3|#4]}
\paragraph{Substitution and holes.}
In \clstr, the structure of the variable context is not a set, but a tree. When substituting
for a variable, we need to replace the variable in the context with the context of the
substituted expression. In general, this can occur anywhere in the tree. To formulate the
statement, we define contexts with holes, written $\Delta[-]$. Note that there is a hole
in the free variable context and in a corresponding part of the coeffect tag:

\begin{equation*}
\begin{array}{rcl}
\Delta[-] &::=& \ctyp{1}{()} \\
       & \sep & \ctyp{r}{(x:\tau)}\\
       & \sep & \ctyp{-}{(-)} \\
       & \sep & \ctyp{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
\end{array}
\end{equation*}
%
Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
general substitution lemma for \clstr.

\begin{figure}

\begin{equation*}
\begin{array}{rcl}
                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
\chole{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \csprd r_2'}{(\Gamma_1', \Gamma_2')}\\
&\multicolumn{2}{l} {
  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
}
\end{array}
\end{equation*}
\caption{The definition of hole filling operation for $\Delta[-]$}
\label{fig:structural-hole-fill}
\vspace{-2em}
\end{figure}

\begin{lemma}[Substitution Lemma]
\label{thm:structural-subst}
If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
then $\chole{r}{\Gamma}{R \csvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
\end{lemma} 
\begin{proof}
Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
$(S, \csvop, 0, \csprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
\end{proof}

\begin{theorem}[Subject reduction]
\label{thm:structural-reduction}
If $\ctyp{r}{\Gamma} \vdash e_1 : \tau$ and $e_1 \twoheadrightarrow_\beta e_2$ then 
$\ctyp{r}{\Gamma} \vdash e_2 : \tau$.
\end{theorem}
\begin{proof}
Direct consequence of Lemma~\ref{thm:structural-subst} (see Appendix~\ref{app:structural-reduction}).
\end{proof}

\paragraph{Local soundness and completeness.}
As with the previous calculus, we want to guarantee that the introduction and
elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
This can be done by showing \emph{local soundness} and \emph{local completeness},
which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
case of subject reduction and the latter is proved by a simple derivation:
%
\begin{theorem}[Local soundness]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
\end{theorem}
\begin{proof}
Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
\end{proof}
%
\begin{theorem}[Local completeness]
If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
\end{theorem}
\begin{proof} The property is proved by the following typing derivation:
%
\begin{equation*}
\inference
  {\inference
     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
     {\ctyp{r \csprd (s \csvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\qedhere
\end{equation*}
\end{proof}
%
In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
guarantees that $s\csvop0 = s$. Recall that in \clflt, the typing derivation for
$\lambda x.f x$ required for local completeness was not the only possible derivation.
In the last step, it was possible to split the coeffect tag arbitrarily between the 
context and the function type.

In the \clstr~calculus, this is not, in general, the case. The $\csprd$ operator is not required 
to be associative and to have units and so a unique splitting may exist.  For example, if we
define $\csprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
$t$, there are unique $r$ and $s$ such that $t = r \csprd s$. However, if the $\csprd$ operation 
has additional properties, then there may be other possible derivation.

@@

\section{Semantics of structural coeffects}
\label{sec:structural-semantics}

The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.

>>>>

\chapter{Flat coeffect language} 
\label{ch:flat}

@@

% comonadic semantics
\newcommand{\iunit}{\mathsf{e}}
\newcommand{\imult}{\oplus}
\newcommand{\iftor}[1]{C^{#1}}
\newcommand{\iftorhat}[1]{\hat{C}^{#1}}

\newcommand{\cosem}[1]{\llbracket #1 \rrbracket}
\newcommand{\interp}[1]{\llbracket #1 \rrbracket}
\newcommand{\ccat}[0]{\mathcal{C}}
\newcommand{\obj}[1]{\textnormal{obj}(#1)}

\newcommand{\cobind}[2]{#1^\dagger_{#2}}
\newcommand{\cmerge}[0]{ \ident{m} }
\newcommand{\csplit}[0]{ \ident{n} }
\newcommand{\counit}[0]{ \varepsilon }
% identity function

% structural coeffect calculus - operation names

\todo{Update the introduction \& edit the whole chapter -- currently, this is just a direct copy
from the accepted ICALP paper on flat coeffects.}

\todo{Other extensions -- add more examples of indexed comonoads, consider recursion and 
fixed-point operator, possibly also consider partially defined operations, which lets us
express type systems that rule out some computations (rather than just static analyses). }

Understanding how programs affect their environment is a well studied area: \emph{effect 
systems}~\cite{effects-talpin-et-al} provide a static analysis of effects and 
\emph{monads}~\cite{monad-notions} provide a unified semantics to different notions of effect.  
Wadler and Thiemann unify the two approaches~\cite{monads-effects-marriage}, \emph{indexing}
a monad with effect information, and showing that the propagation of effects in an effect system 
matches the semantic propagation of effects in the monadic approach.

No such unified mechanism exists for tracking the context requirements. We use the term 
\emph{coeffect} for such contextual program properties. 
Notions of context have been previously captured using comonads
\cite{comonads-notions} (the dual of monads) 
and by languages derived from modal logic
\cite{logic-modal-reconstruction,logic-cmtt}, but these approaches
do not capture many useful examples which motivate our work. 
We build mainly on the former comonadic direction (\S\ref{sec:comonads}) and discuss the
modal logic approach later (\S\ref{sec:related}).

We extend a simply typed lambda calculus with a coeffect system based on comonads,
replicating the successful approach of effect systems and monads.

\paragraph{Examples of coeffects.}
We present three examples that do not fit the traditional approach of
\emph{effect systems} and have not been considered using the \emph{modal logic}
perspective, but can be captured as \emph{coeffect systems} (\S\ref{sec:motivation}) --
the tracking of implicit dynamically-scoped parameters (or resources),
analysis of variable liveness, and tracking the
number of required past values in dataflow computations.

\paragraph{Coeffect calculus.}
Informed by the examples, we identify a general algebraic structure for coeffects. 
From this, we define a general \emph{coeffect calculus} that unifies the motivating 
examples (\S\ref{sec:calculus}) and discuss its syntactic properties (\S\ref{sec:syntactic}).

\paragraph{Indexed comonads.}
Our categorical semantics (\S\ref{sec:comonads}) extends the work of Uustalu and Vene 
\cite{comonads-notions}. By adding annotations, we generalize comonads to \emph{indexed comonads}, 
which capture notions of computation not captured by ordinary comonads.

\section{Motivation}
\label{sec:motivation}

Effect systems, introduced by Gifford and Lucassen
\cite{effects-gifford}, track \emph{effects} of computations, such as
memory access or message-based
communication~\cite{effects-messagepassing}. Their approach augments
typing judgments with effect information: $\Gamma \vdash e :
\tau, F$.  Wadler and Thiemann explain how this shapes
effect analysis of lambda abstraction~\cite{monads-effects-marriage}:
%
\begin{quote}
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the body.}
\end{quote}
%
In contrast to the static analysis of \emph{effects}, the analysis of \emph{context-dependence} 
does not match this pattern. In the systems we consider, lambda abstraction places 
requirements on both the \emph{call-site} (latent requirements) and the \emph{declaration-site} 
(immediate requirements), resulting in different syntactic properties. 
We informally discuss three examples first that demonstrate how contextual
requirements propagate. Section (\S\ref{sec:calculus}) then unifies these 
in a single calculus.

We write coeffect judgements $\ctyp{s}{\Gamma} \vdash e : \tau$ where the 
coeffect annotation $s$ associates context requirements with the 
free-variable context $\Gamma$. Function types have the form $\ctyp{s}{\tau_1} \rightarrow \tau_2$ associating
\emph{latent} coeffects $s$ with the parameter.
The $\ctyp{s}{\Gamma}$ syntax and $\ctyp{s}{\tau}$ types are a result of the
indexed comonadic semantics (\S\ref{sec:comonads}).

% --------------------------------------------------------------------------------------------------
\vspace{-1em}
\subsection{Implicit parameters and resources.}
\label{sec:coeffects-res}

Implicit parameters~\cite{app-implicit-parameters} are \emph{dynamically-scoped} variables.
They can be used to parameterize a computation without propagating arguments explicitly through a 
chain of calls and are part of the context in which expressions evaluate. As correctly expected
\cite{app-implicit-parameters}, they can be modelled by comonads. Rebindable resources in distributed
computations follow a similar pattern, but we discuss implicit parameters for simplicity.

The following function prints a number using implicit parameters
\ident{?culture} (determining the decimal mark) and \ident{?format}
(the number of decimal places):
%
\begin{equation*}
\lambda n . \ident{printNumber}~n~\ident{?culture}~\ident{?format}
\end{equation*}
%
Figure~\ref{fig:example-resources} shows a type-and-coeffect system
tracking the set of an expression's implicit parameters. 
For simplicity here, all implicit parameters have type $\rho$.

Context requirements are created in (\emph{access}), while (\emph{var}) requires no
implicit parameters; (\emph{app}) combines requirements of
both sub-expressions as well as the latent requirements of the
function. The (\emph{abs}) rule is where the example differs from
effect systems.  Function bodies can access the union of the
parameters (or resources) available at the declaration-site ($\ctyp{r}{\Gamma}$) and
at the call-site ($\ctyp{s}{\tau_1}$). Two of the nine permissible
judgements for the above example are:
%
\newcommand{\cfebody}{(\ldots)} % \ident{countFutureEvents}
\begin{equation*}
\begin{array}{rcl}
\ctyp{\emptyset}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?culture}, \ident{?format} \}}{\ident{int}} \rightarrow \ident{string}\\
\ctyp{\{\ident{?culture}, \ident{?format}\}}{\Gamma} &\vdash& \cfebody : \ctyp{\{\ident{?format}\}}{\ident{int}} \rightarrow \ident{string}
\end{array}
\end{equation*}
%
The coeffect system infers multiple, \ie{} non-principal, coeffects for functions.
Different judgments are desirable depending on how a function is used. 
In the first case, both parameters have to be provided by the caller.
In the second, both are available at declaration-site, but \ident{?format} may be
rebound (precise meaning is provided by the monoidal structure on the product comonad in \S\ref{sec:comonads}).

Implicit parameters can be captured by the
\emph{reader} monad, where parameters are associated with the function codomain $\mtyp{\emptyset}{(\ident{int} \rightarrow
  \mtyp{\{\ident{?culture}, \ident{?format} \}}{\ident{string}})}$,
modelling only the first case. Whilst the reader monad can be extended to model rebinding, 
the next example cannot be structured by \emph{any} monad.

@@

\begin{figure}[t]
\vspace{-1.4em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{\emptyset}{\Gamma} \vdash x : \tau }
%
\quad\quad
%
\inference[(\emph{app})\;]
  {\ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 &
   \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 }
  {\ctyp{r \cup s \cup t}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\vspace{-0.7em}
\begin{equation*}
\inference[(\emph{access})\;]
  {}
  {\ctyp{ \{\ident{?a} \} }{\Gamma} \vdash \ident{?a} : \rho }
%
\quad\quad\quad
%
\inference[(\emph{abs})\;]
  {\ctyp{r \cup s}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{r}{\Gamma} \vdash \lambda x.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\end{equation*}
\vspace{-0.9em}
\caption{Selected coeffect rules for implicit parameters}
\label{fig:example-resources}
\vspace{-1em}
\end{figure}

@@

\begin{figure}[!b]
\vspace{-2em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{\ident L}{\Gamma} \vdash x : \tau }
\quad
\inference[(\emph{app})\;]
  { \ctyp{s}{\Gamma} \vdash e_2 : \tau_1 &
    \ctyp{r}{\Gamma} \vdash e_1 : \ctyp{t}{\tau_1} \rightarrow \tau_2 }
  {\ctyp{r \sqcup (s \sqcap t)}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\vspace{-1.5em}
\caption{Selected coeffect rules for liveness analysis}
\label{fig:example-need}
\vspace{-1.2em}
\end{figure}

@@

\vspace{-1em}
\subsection{Liveness analysis.}

Liveness analysis detects whether a free variable of an expression may be used (\emph{live}) or 
whether it is definitely not needed (\emph{dead}). A compiler can remove
bindings to dead variables as the result is never used.

We start with a restricted analysis and briefly mention how to make it practical
later (\S\ref{sec:related}). The restricted form is interesting theoretically as it gives rise
to the indexed Maybe comonad (\S\ref{sec:comonads}), which is a basic but instructive example.

A coeffect system in Fig.~\ref{fig:example-need} detects whether all variables are dead 
($\ctyp{\ident D}{\Gamma}$) or whether at least one variable is live ($\ctyp{\ident L}{\Gamma}$). 
Variable access (\emph{var}) is annotated with \ident L and constant access with \ident D. That is,
if $c\in\mathbb{N}$ then $\ctyp{\ident D}{\Gamma} \vdash c : \ident{int}$.
A dead context may be marked as live by
letting $\ident D \sqsubseteq \ident L$ and adding sub-coeffecting (\S\ref{sec:calculus}).

The (\emph{app}) rule is best understood by discussing its semantics. Consider first 
\emph{sequential composition} of (semantic) functions $g, f$ annotated with $r, s$. 
The argument of $g \circ f$ is live only when arguments of both $f$ and $g$ are live. 
The coeffect semantics captures the additional behaviour that $f$ is not evaluated when
$g$ ignores its input (regardless of the evaluation order of the underlying language).
We write $r \sqcap s$ for a conjunction (returning \ident L iff $r = s = \ident{L}$). 
Secondly, a \emph{pointwise composition} passes the same
argument to $g$ and $h$. The parameter is live if either the parameter
of $g$ or $h$ is live ($r \sqcup s$). Application combines the
two operations, so the context $\Gamma$ is live if it is needed by
$e_1$ \emph{or} by the function value \emph{and} by $e_2$.

An (\emph{abs}) rule (not shown) compatible with the structure in Fig.~\ref{fig:example-resources}
combines the context annotations using $\sqcap$. Thus, if the body uses some 
variables, both the function argument and the context of the declaration-site are marked as live.

Liveness cannot be modelled using monads as $\tau_1 \rightarrow \mtyp{r}{\tau_2}$.
In call-by-value languages, the argument $\tau_1$ is always evaluated. Using 
indexed comonads (\S\ref{sec:comonads}), we
model liveness as $\ctyp{r}{\tau_1} \rightarrow \tau_2$ where
$\ctyp{r}$ is the parametric type $\mathsf{Maybe}~\tau = \tau + 1$
(which contains a value $\tau$ when $r=\ident{L}$ and does not contain
value when $r=\ident{D}$).

@@

\begin{figure}[t]
\vspace{-1em}
\begin{equation*}
\inference[(\emph{var})\;]
  {x : \tau \in \Gamma}
  {\ctyp{0}{\Gamma} \vdash x : \tau }
%
\quad\quad
% 
\inference[(\emph{app})\;]
  {\ctyp{m}{\Gamma} \vdash e_1 : \ctyp{p}{\tau_1} \rightarrow \tau_2 &
   \ctyp{n}{\Gamma} \vdash e_2 : \tau_1 }
  {\ctyp{\textit{max}(m, n + p)}{\Gamma} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\inference[(\emph{prev})\;]
  {\ctyp{n}{\Gamma} \vdash e : \tau}
  {\ctyp{n+1}{\Gamma} \vdash \kvd{prev}~e : \tau}
%
\quad\quad
%
\inference[(\emph{abs})\;]
  {\ctyp{\textit{min}(m, n)}{(\Gamma, x:\tau_1)} \vdash e : \tau_2}
  {\ctyp{m}{\Gamma} \vdash \lambda x.e : \ctyp{n}{\tau_1} \rightarrow \tau_2 }
\end{equation*}
\vspace{-1em}
\caption{Selected coeffect rules for causal data flow}
\label{fig:example-dataflow}
\vspace{-1em}
\end{figure}

\vspace{-1em}
\subsection{Efficient dataflow.}
Dataflow languages (\eg~\cite{app-lucid}) declaratively describe
computations over streams. In \emph{causal} data flow, program may
access past values -- in this setting, a function $\tau_1 \rightarrow
\tau_2$ becomes a function from a list of historical values $[\tau_1]
\rightarrow \tau_2$. A coeffect system here tracks how many past values
to cache.

Figure~\ref{fig:example-dataflow} annotates contexts with an integer specifying the maximum
number of required past values. The current value is always present, so (\emph{var}) is annotated with $0$.
The expression $\kvd{prev}~e$ gets the previous value of stream $e$ and requires one
additional past value (\emph{prev}); \eg~$\kvd{prev}~(\kvd{prev}~e)$ requires 2 past values.

The (\emph{app}) rule follows the same intuition as for liveness. Sequential composition
adds the tags (the first function needs $n + p$ past values to produce $p$ past inputs
for the second function); passing the context to two subcomputations requires the maximum
number of the elements required by the two subcomputations. The (\emph{abs}) rule for data-flow 
needs a distinct operator -- \emph{min} -- therefore, the declaration-site and 
call-site must each provide at least the number of past values required by the function body
(the body may use variables coming from the declaration-site as well as the argument).

The soundness follows from our categorical model (\S\ref{sec:comonads}). Uustalu and Vene
\cite{comonads-notions} model causal dataflow computations using a non-empty list comonad 
$\ident{NeList}~\tau = \tau \times (\ident{NeList}~\tau + 1)$. However, such model leads to
(inefficient) unbounded lists of past elements. The above static analysis provides an
approximation of the number of required past elements and so we use just fixed-length lists.

@@

\vspace{-1em}
\subsection{Coeffect typing rules.}
Figure~\ref{fig:coeffect-typing} shows the
rules of the coeffect calculus, given some coeffect algebra $(S, \imult, \vee, \wedge, \iunit)$.
The context required by a variable access (\emph{var}) is annotated with $\iunit$.
The sub-coeffecting rule (\emph{sub}) allows the contextual requirements of an expression 
to be generalized.
